---
output: 
  github_document:
    html_preview: false
knit: rmarkdown::render
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


[![R-CMD-check](https://github.com/Chukyhenry/PosiR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Chukyhenry/PosiR/actions)

**PosiR** provides tools for post-selection inference (PoSI) in linear regression models. Post-Selection Inference addresses the challenge of performing valid statistical inference after model selection, ensuring that confidence intervals maintain their nominal coverage probability (e.g., 95%) even when the model is chosen based on the data. The package implements simultaneous confidence intervals using bootstrap-based max-t statistics, following Algorithm 1 from Kuchibhotla, Kolassa, and Kuffner (2022).

## Installation
You can install the development version of `PosiR` from GitHub:
```{r eval = FALSE}
# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
# Install PosiR
devtools::install()

# Optional dependencies for vignette and examples
install.packages(c("dplyr", "pbapply"))
```

## Example: Simultaneous Confidence Intervals

This example demonstrates how to use `simultaneous_ci()` to compute simultaneous confidence intervals for regression coefficients across a set of models:

```{r example}
library(PosiR)

# Simulate data
set.seed(123)
X <- matrix(rnorm(100 * 3), 100, 3)
colnames(X) <- c("X1", "X2", "X3")
y <- 1 + X[, "X1"] * 0.5 + rnorm(100)  # True intercept = 1, X1 coefficient = 0.5

# Define model universe (column indices of X)
Q <- list(
  model1 = 1:2,  # Model with X1, X2
  model2 = 1:3   # Model with X1, X2, X3
)

# Compute simultaneous confidence intervals
result <- simultaneous_ci(X, y, Q, B = 500, verbose = FALSE)

# View results
print(result$intervals)

# Plot the intervals
plot(result, main = "Simultaneous Confidence Intervals", las.labels = 1)

```
## Interpretation

The output `result$intervals` provides the coefficient estimates and simultaneous 95% confidence intervals for each model in `Q`. For example:

The `(Intercept)` and `X1` intervals in `model1` should contain their true values (1 and 0.5, respectively).

The intervals are wider than naive intervals to account for model selection uncertainty, ensuring valid coverage across all models in `Q`.

## Learn More
Vignette: Run **vignette("Vignette")**.

Source Paper: Kuchibhotla, A., Kolassa, J., & Kuffner, T. (2022). Post-selection inference. Annual Review of Statistics and Its Application, 9(1), 505–527. DOI: 10.1146/annurev-statistics-100421-044639.
